---
title: "Stock Exchange Data Analysis"
author: "Vladyslav Karamushka"
date: "2026-04-27"
output: 
  html_document:
    code_folding: hide
---

## Dataset Overview

The analysis is based on daily trading data sourced from Yahoo Finance, specifically targeting companies within the S&P 500 (US) and STOXX Europe 600 (EU) indices, which was created by sampling tickers and taking data only for those companies to get symmetrical sector coverage. The dataset consists of **54,015 daily observations** across 15 variables, split relatively evenly between the US (27,709 observations) and the EU (26,306 observations). It comprises **121 matched companies** across the US and EU with symmetrical sector coverage. Minor differences in observation counts per company (capped at a maximum variance of 20 days) are statistically insignificant and will not affect the analysis results. All qualitative variables have been encoded as factors to facilitate structured, level-based statistical modeling.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE
)

library(dplyr)
library(ggplot2)
library(ggdist)
library(nortest)
library(knitr)
library(tidyverse)

long_dataset <- read.csv("long_dataset.csv")
```

```{r sampling, include=FALSE}

df_grp_ticker = long_dataset %>% distinct(ticker, region, sector) 
df_grp_region = df_grp_ticker %>% group_by(region) %>% summarise(count = n(), .groups = 'drop')

df_grp_sector = df_grp_ticker %>% group_by(region,sector) %>% summarize(count=n(), .groups = 'drop')

valid_sectors = df_grp_ticker %>% filter(region %in% c("EU","US")) %>% group_by(sector) %>%
  summarise(n_regions = n_distinct(region), .groups = 'drop') %>% filter(n_regions==2)

#after filtration
df_filtered = df_grp_ticker %>% filter(sector %in% valid_sectors$sector)
df_filtered_region = df_filtered %>% group_by(region) %>% summarize(count=n(), .groups =  'drop')

df_filtered_sector = df_filtered %>% group_by(region, sector) %>% summarize(count = n(), .groups =  'drop')



#sampling
set.seed(42)
n_sample <- 121
# --- Compute allocation from EU counts ---
eu_sector_counts <- df_filtered_sector %>%
  filter(region == "EU")

allocation <- eu_sector_counts %>%
  mutate(
    raw       = count / sum(count) * n_sample,
    floor_n   = floor(raw),
    remainder = raw - floor_n
  ) %>%
  arrange(desc(remainder)) %>%
  mutate(
    extra   = row_number() <= (n_sample - sum(floor_n)),
    alloc_n = floor_n + as.integer(extra)
  ) %>%
  select(sector, alloc_n)

# --- Sample US tickers per sector ---
sampled_tickers <- df_filtered %>%
  filter(region == "US") %>%
  left_join(allocation, by = "sector") %>%
  group_by(sector) %>%
  group_modify(~ slice_sample(.x, n = unique(.x$alloc_n))) %>%
  ungroup() %>%
  select(ticker, sector)


write.csv(sampled_tickers, "sampled_tickers.csv")

long_dataset <- long_dataset %>% filter(region == "EU" | ticker %in% sampled_tickers$ticker)
long_dataset <- long_dataset[long_dataset$sector != "cash and/or derivatives", ]
```

```{r sampling2, include=FALSE}
long_dataset$cap_group <- as.factor(long_dataset$cap_group)
long_dataset$region <- as.factor(long_dataset$region)
long_dataset$sector <- as.factor(long_dataset$sector)
```

## Distributional Assumptions & Statistical Testing

```{r return qqplot}
qqnorm(long_dataset$return, main = "Return Q-Q plot")
```

```{r log volatility qqplot}
qqnorm(log(long_dataset[long_dataset$volatility > 0, "volatility"]), main = "Volatility Q-Q plot")
```

```{r lillieforse}
lillie.test(long_dataset$volatility[1:nrow(long_dataset)*20])
```

Initial hypotheses assuming a normal distribution for key financial metrics (Returns and Volatility) were evaluated and subsequently rejected. Quantitative assessments, including Kolmogorov-Smirnov tests and Q-Q plot visualizations, confirm a definitive lack of normality across the dataset.

Crucially, the return data exhibits "fat tails" property. This structural characteristic indicates that extreme market movements (both positive and negative) occur with a significantly higher probability than a standard normal distribution would predict. Due to this non-normality, non-parametric statistical methods were prioritized for subsequent comparative analysis. 

## Sectoral Return Analysis & Outlier Impact

To evaluate the "typical" baseline performance across sectors, preliminary visualizations (violin and box plots) were constructed using a filtered dataset. Values being further from 0 than for 0.03 were cut to preserve plot readability because of extreme values existence.

```{r cutting outliers, echo=FALSE, message=FALSE, warning=FALSE}
q1 <- quantile(long_dataset$return, .25)
q3 <- quantile(long_dataset$return, .75)
blow <- q1 - 1.5 * (q3-q1) 
bhigh <- q3 + 1.5 * (q3-q1)
```


```{r violin sectors}
ggplot(long_dataset[long_dataset$return >= blow & long_dataset$return <= bhigh, ],
       aes(x = sector, y = return, fill = sector)) +
  geom_violin() +
  geom_boxplot(width = 0.15, fill = "white") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  coord_cartesian(ylim = c(-0.03, 0.03))
```

* **Baseline Performance:** The filtered data indicates that the Communication sector has the highest average returns, along with substantial instability. Healthcare presents an optimal risk-reward profile, offering profitability comparable to Energy and Tech, but with higher stability. Utilities represent the most stable sector overall, while Consumer Goods and Finance represent the least profitable segments.

* **The Impact of Tail Events:** Reintegrating the excluded outliers (approximately 8% of observations) significantly alters the sectoral profile, highlighting the danger of relying solely on filtered data. When accounting for extreme trading days, the mean returns for Healthcare and Energy outpace the rest of the market by a factor of 1.5.

* **Risk Symmetry:** Significant divergences between mean and median returns -— particularly in the Communication and Energy sectors —- underscore high vulnerability to asymmetric risk. For example, while Energy frequently captures outsized positive returns, Communication is uniquely vulnerable to rare but severe drawdowns.

## Regional Volatility Dynamics

```{r trimming, echo=FALSE, message=FALSE, warning=FALSE}
threshold_us <- quantile(long_dataset$volatility[long_dataset$region == "US"], 0.99, na.rm = TRUE)
threshold_eu <- quantile(long_dataset$volatility[long_dataset$region == "EU"], 0.99, na.rm = TRUE)
data_vol_cut <- long_dataset %>%
  group_by(region) %>%
  filter(
    (region == "US" & volatility <= threshold_us) |
    (region == "EU" & volatility <= threshold_eu)
  ) %>%
  ungroup()
```

```{r raincloud}
ggplot(data_vol_cut, aes(x = region, y = volatility, fill = region, color = region)) +
  
  stat_halfeye(
    adjust = 0.5, 
    width = 0.6, 
    .width = 0, 
    justification = -0.2, 
    point_colour = NA, 
    alpha = 0.7
  ) +
  
  geom_boxplot(
    width = 0.15, 
    outlier.shape = NA, 
    alpha = 0.5, 
    color = "black"
  ) +
  
  coord_flip() + 
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Volatility distribution (trimmed by 99'th quantile)",
    x = "Region", 
    y = "Volatility"
  )
```

Regional volatility profiles were analyzed utilizing raincloud plots to map data density and distribution shape. Because of high number of observations and density of points under the plot, the "rain" was not been added. While baseline median volatility appears visually comparable between the two groups, advanced statistical testing reveals fundamental divergences.

Given the non-normal distribution of the volatility data (and its logarithmic transformations), a Mann-Whitney U test was applied to compare regional medians.

```{r mann-whitney}
idx <- seq(from = 20, to = nrow(long_dataset), by = 20) 

small_data <- long_dataset[idx, ]

wilcox.test(volatility ~ region, 
            data = small_data, 
            subset = region %in% c("US", "EU"))
```

The resulting p-value confirms a statistically significant difference in baseline volatility between the regions. Overall, the EU market exhibits a structurally higher mean and variance, driven largely by a higher frequency of extreme volatility outliers compared to the US market.

```{r statistics for vol by region}
summary_stats_volatility <- long_dataset %>%
  group_by(region) %>%
  summarise(
    mean_ret = mean(volatility, na.rm = TRUE),
    median_ret = median(volatility, na.rm = TRUE),
    sd_ret = sd(volatility, na.rm = TRUE),     
    n_obs = n()                                
  )

print(summary_stats_volatility)
```

## Cross-Sectional Volatility Matrix

To present in a clear way the interaction between geography and industry, a median-based volatility heatmap was developed. Utilizing medians provides a clearer picture of structural volatility by ignoring the distortion caused by extreme tail events.

```{r median volatility heatmap}
heatmap_data <- long_dataset %>%
  group_by(region, sector) %>%
  summarise(median_vol = median(volatility, na.rm = TRUE), .groups = "drop")

ggplot(heatmap_data, aes(x = region, y = sector, fill = median_vol)) +
  geom_tile(color = "white") +
  
  geom_text(aes(label = round(median_vol, 4)), color = "black", size = 3) +
  
  scale_fill_gradient(low = "white", high = "steelblue", name = "Volatility") +
  
  theme_minimal() +
  labs(
    title = "Volatility median heatmap",
    x = "Sector",
    y = "Region"
  ) +
  theme(panel.grid = element_blank())
```

* **Sectoral Extremes:** The multidimensional analysis reveals that the Technology sector is definitively the most volatile segment across the entire dataset. This refines our earlier aggregate observations (which pointed to Energy as the most unstable), proving that standard sector behavior is not representative for every company and there are many extreme cases which is hard to model.

* **Regional Disparities:** While 4 out of 10 sectors display marginally higher median volatility in the US, the magnitude of regional divergence differs in specific industries. For instance, the Energy and Technology sectors are substantially more volatile in Europe than their American counterparts.

## Volatility-Volume Dynamics & Market Liquidity

The next phase of the analysis investigates the relationship between trading volume and price volatility. While both metrics share non-negative structural boundaries, initial hypotheses suggesting a strong positive correlation were largely unsupported by the data.

To effectively map this relationship and mitigate visual distortion from extreme tail events, the analytical domain was bounded (Volumes up to 10,000,000 shares; Volatility up to 0.10).

```{r cor plot}
ggplot(data_vol_cut, aes(x = volume, y = volatility)) +
  geom_point(alpha = 0.3, size = 0.5) + 

  geom_smooth(method = "lm", color = "red", linetype = "dashed") + 
  
  coord_cartesian(xlim = c(0, 10^7), ylim = c(0, 0.1)) +
  
  theme_minimal() +
  labs(title = "Correlation vizualisation",
       subtitle = "Dashed line shows general trend")
```

**Key Findings:**

* **Weak Linear Correlation:** The aggregate data demonstrates a structurally weak linear correlation between trading volume and volatility, approaching zero. Due to the high frequency of outliers, traditional linear regression models proved insufficient for capturing the true dynamics between these variables.

* **Pronounced Heteroscedasticity:** The most critical insight from this bivariate analysis is the presence of severe heteroscedasticity. Specifically, the variance (or spread) of volatility compresses significantly as trading volume increases.

* **Liquidity and Market Vulnerability:** This narrowing variance is a direct reflection of market liquidity mechanics. While the dataset naturally contains a higher concentration of lower-volume equities, a clear structural trend emerges:

* **Low-Volume Dynamics (Baseline Stability vs. Tail Risk):** For the majority of lower-volume equities, general market interest is subdued. Because their price movements have a marginally smaller impact on broader market capitalization, their day-to-day baseline volatility remains relatively low. However, this same lack of liquidity creates severe vulnerability to tail events, resulting in extreme, though infrequent, volatility spikes (outliers).

* **High-Volume Dynamics (Elevated Baseline vs. Shock Absorption):** Conversely, high-volume equities —- typically large firms —- experience a higher baseline level of volatility. This is driven by intense, continuous market interest, and the major impact these equities have on overall index valuations. Yet, while their day-to-day fluctuations are greater, it requires substantially more capital to distort their prices, lowering the risk of the extreme outlier events seen in thinly traded stocks.

```{r spearman}
cor.test(data_vol_cut$volatility, data_vol_cut$volume, method = "spearman")
```

Small, but statistically significant correlation existence confirms the result of the Spearman's rank sum test.

## Predictive Modeling: Volatility Regression Analysis

Financial markets are inherently chaotic, the direct prediction of raw daily financial variables is highly complex and typically requiring advanced proprietary modeling. Therefore, the objective of this linear regression model is not pure forecasting, but rather quantifying the structural impact and statistical significance of various market factors on asset stability.

**Data Transformation:**

```{r logaritmizing}
long_dataset$log_volatility <- log(long_dataset$volatility)
long_dataset_clear <- long_dataset %>% filter(!is.na(log_volatility) & !is.infinite(log_volatility))
```

```{r clearing}
long_dataset_clear$log_volume <- log(long_dataset_clear$volume)
long_dataset_clear <- long_dataset_clear %>% filter(!is.na(log_volume) & !is.infinite(log_volume))
```

```{r logarithmized volume distribution plot}
ggplot(long_dataset_clear, aes(x = log_volume)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 100, fill = "steelblue", color = "white", alpha = 0.7) +
  theme_minimal()
```

```{r trimming for model}
long_dataset_clear_cut <- long_dataset_clear %>% 
  filter(
    log_volume >= quantile(log_volume, 0.01, na.rm = TRUE),
    log_volume <= quantile(log_volume, 0.99, na.rm = TRUE),
    
    log_volatility >= quantile(log_volatility, 0.01, na.rm = TRUE),
    log_volatility <= quantile(log_volatility, 0.99, na.rm = TRUE),
    
    log_market_cap >= quantile(log_market_cap, 0.01, na.rm= TRUE),
    log_market_cap <= quantile(log_market_cap, 0.99, na.rm = TRUE)
  )
```

* **Logarithmic Scaling & Outlier Management:** To address severe right-skewness and normalize the distributions for linear modeling, logarithmic transformations were applied to both Volatility and Volume (Market Capitalization was previously transformed). Extreme outliers were systematically trimmed to prevent the distortion of regression coefficients.

```{r lagging variable}
long_dataset_for_lm <- long_dataset_clear_cut %>%
  arrange(region) %>%
  group_by(ticker) %>%
  mutate(log_prev_volatility = lag(log_volatility)) %>%
  filter(!is.na(log_prev_volatility)) %>%
  ungroup()

ggplot(long_dataset_for_lm, aes(x = log_prev_volatility, y = log_volatility)) + geom_point(alpha = 0.3, size =   0.5) + geom_smooth(method = "lm", color = "red", linetype = "dashed") + 
  theme_minimal() +
  labs(title = "Correlation visualisation for previous observations",
       subtitle = "Dashed line shows general trend")
```

* **Mitigating Autocorrelation:** Financial volatility exhibits a strong "memory" effect (autocorrelation) and heteroscedasticity. Current volatility is deeply correlated with historical observations. To neutralize error autocorrelation and capture this dynamic, a lagged variable (log_prev_volatility, representing the previous day's volatility) was introduced. Raw open/close prices and baseline returns were excluded from the predictor set due to their instability and mathematical redundancy with the target variable.

## Regression Results & Key Market Insights
 
```{r model}
model <- lm(log_volatility ~ log_prev_volatility + log_market_cap + region + sector + log_volume, data = long_dataset_for_lm)

summary(model)
```

The linear regression model has an exceptionally high Adjusted R-squared value approaching 1.0. While absolute predictive perfection is atypical in financial modeling, this result validates several critical market mechanics:

* **The Volatility Clustering Phenomenon:** The lagged volatility variable (log_prev_volatility) accounts for approximately 98% of the model's explanatory power. This quantitatively confirms the market phenomenon known as volatility clustering —- periods of high market turbulence are statistically likely to be followed by continued high turbulence, and tranquil periods tend to persist.

* **Size vs. Liquidity (Market Cap & Volume):** Controlling for lagged volatility, Market Capitalization exerts a negative impact on volatility, confirming that larger, mega-cap companies are structurally more stable and resistant to price shocks. Conversely, Trading Volume exhibits a positive coefficient, aligning with our earlier findings regarding the baseline instability of highly traded, high-interest equities.

* **Regional Stability:** The regression confirms regional differences, proving with statistical significance that US equities maintain a lower baseline volatility compared to the EU cohort, holding other factors constant.

* **Sectoral Anomalies & Validations:** The model definitively identifies the Technology sector as the most volatile segment in a "standard case," aligning with two out of three of our preceding exploratory analyses. Interestingly, the regression suggested high volatility in the Healthcare sector, contradicting our earlier baseline findings. However, given the high p-values associated with this specific sector coefficient, this is likely a statistical artifact driven by residual error autocorrelation rather than a true structural shift, and should not overrule the primary exploratory data.
